Very simplified likelihood

### 2024-02-12 inst/very_simplified_likelihood.Rmd
typo
typos

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
An observation is indexed by $r$, the number of $b$ clones finishing
ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
strength is $a+nb=n+(n-1)a$.

@@

@@

```{r}
f_single <- function(a,r,n,log=FALSE){  # not vectorised
  B <- 1/(1-a)
  if(log){
     if(r==n){  # last place
       out <- lgamma(B)-lgamma(B+n)
     } else {   # not last
       out <- log(B-1) - sum(log(B + B+(n-r-1):(n-1)))
     }
   } else {  # not-log 
     if(r==n){
      out <- gamma(B)/gamma(B+n)
     } else {
       out <- (B-1)/prod(B+(n-r-1):(n-1))
     }
   }	
   return(out)
}

@@

but this is incorrect: there are two clones of (strength) $b$, and the
likelihood arguments above do not distinguish between their finishing
order.  We can name the competitors $a$, $b_1$ and $b_2$ and specify
that $b_1$ and $b_2$ both have strength $b$.  Then "coming second"
means that the finishing order was $b_1,a,b_2$ or $b_2,a,b_1$.  Noting
that these events are disjoint, the probability would be

\[\mbox{prob of coming second}   =
\operatorname{P}(b_1,a,b_2)      +
\operatorname{P}(b_2,a,b_1)      =
\frac{b}{a+2b}\cdot\frac{a}{a+b} +
\frac{b}{a+2b}\cdot\frac{a}{a+b} = \frac{2(1-b)}{1+b}
\]

Similarly for the focal competitor coming last we need to sum the
probabilities of finishing orders $b_1,b_2,a$ and $b_2,b_1,a$:

\[\mbox{prob of coming third}=
\operatorname{P}(b_1,b_2,a) +
\operatorname{P}(b_2,b_1,a) =
\frac{b}{a+2b}\cdot\frac{b}{a+b}+
\frac{b}{a+2b}\cdot\frac{b}{a+b}=
\frac{2b^2}{1+b}.
\]

@@

\[
\mbox{prob of coming first}      =
\operatorname{P}(a,b_1,b_2)      +
\operatorname{P}(a,b_2,b_1)      =
\frac{b}{a+2b}\cdot\frac{b}{b+b} +
\frac{b}{a+2b}\cdot\frac{b}{b+b} = \frac{1-b}{1+b}
\]

@@

```{r,label=likevis,cache=TRUE}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 7
jj <- f(a,3,n,log=FALSE)
plot(a,jj/max(jj,na.rm=TRUE),type='n')
grid()
for(r in seq(from=0,to=n)){
  y <- f_vec_a(a,r,n,log=FALSE)
  if(r==0){y[length(y)] <- 1}  
  y <- y/max(y,na.rm=TRUE)
  if(r>0){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rainbow(n+1)[r+1])
  segments(x0=r/(n),y0=0.9,y1=1)
  text(x=r/(n),y=0.8,r)
}
}
plotter()
pdf(file="ninelikes.pdf")
plotter()
dev.off()
```

@@

```{r calculatelikes,cache=TRUE}
a <- seq(from=0.15,to=0.5,by=0.01)
La <- f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)
La <- La - max(La,na.rm=TRUE)
```

@@

```{r calculatelikes1,cache=TRUE}
a <- seq(from=0.005,to=0.6,by=0.005)
La <- f_vec_arn(a,r=results$place[1]-1,results$runners[1]-1,log=TRUE)
La <- La - max(La,na.rm=TRUE)
plot(a,La,type='b')
abline(h=c(0,-2))
```

@@

```{r,label=calcbothlikes,cache=FALSE}
x <- seq(from=0.1,to=0.8,by=0.01)
jj <- as.matrix(expand.grid(x,x))
L <- apply(jj,1,supp2,results$place,results$runners)
L <- L - max(L,na.rm=TRUE)
L <- matrix(L,length(x),length(x))
```

@@

```{r,cache=TRUE}
x <- seq(from=0.1,to=0.95,by=0.01)
jj <- as.matrix(expand.grid(x,x))
L <- apply(jj,1,supp2_ed,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)
L <- L - max(L,na.rm=TRUE)
L <- matrix(L,length(x),length(x))
```

>>>>

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
An observation is indexed by $r$, the number of $b$ clones finishing
ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
strength is $a+nb=1+b(n-1)$.

@@

This is

\[
\frac{(B-1)\Gamma(B+n-r-1)}{\Gamma(B+n)}
\]

@@

```{r}
f_single <- function(a,r,n,log=FALSE){  # not vectorised
  B <- 1/(1-a)
  if(log){
     if(r==n){  # last place
       out <- lgamma(B)-lgamma(B+n)
     } else {   # not last
       out <- log(B-1) - sum(log(B + (n-r-1):(n-1)))
     }
   } else {  # not-log 
     if(r==n){
      out <- gamma(B)/gamma(B+n)
     } else {
       out <- (B-1)/prod(B+(n-r-1):(n-1))
     }
   }	
   return(out)
}

@@

but this is incorrect: there are two clones of (strength) $b$, and the
likelihood arguments above do not distinguish between their finishing
order.  We can name the competitors $a$, $b_1$ and $b_2$ and specify
that $b_1$ and $b_2$ both have strength $b$.  Then "coming second"
means that the finishing order was $b_1\succ a\succ b_2$ or $b_2\succ
a\succ b_1$.  Noting that these events are disjoint, the probability
would be

\[\mbox{prob of coming second}   =
\operatorname{P}(b_1\succ a\succ b_2)      +
\operatorname{P}(b_2\succ a\succ b_1)      =
\frac{b}{a+2b}\cdot\frac{a}{a+b} +
\frac{b}{a+2b}\cdot\frac{a}{a+b} = \frac{2(1-b)}{1+b}
\]

Similarly for the focal competitor coming last we need to sum the
probabilities of finishing orders $b_1\succ b_2\succ a$ and $b_2\succ
b_1\succ a$:

\[\mbox{prob of coming third}=
\operatorname{P}(b_1\succ b_2\succ a) +
\operatorname{P}(b_2\succ b_1\succ a) =
\frac{b}{a+2b}\cdot\frac{b}{a+b}+
\frac{b}{a+2b}\cdot\frac{b}{a+b}=
\frac{2b^2}{1+b}.
\]

@@

\[
\mbox{prob of coming first}      =
\operatorname{P}(a\succ b_1\succ b_2)      +
\operatorname{P}(a\succ b_2\succ b_1)      =
\frac{b}{a+2b}\cdot\frac{b}{b+b} +
\frac{b}{a+2b}\cdot\frac{b}{b+b} = \frac{1-b}{1+b}
\]

@@

```{r,label=likevis,cache=F}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 7
jj <- f_vec_arn(a,3,n,log=FALSE)
plot(a,jj/max(jj,na.rm=TRUE),type='n')
grid()
for(r in seq(from=0,to=n)){
  y <- f_vec_a(a,r,n,log=FALSE)
  if(r==0){y[length(y)] <- 1}  
  y <- y/max(y,na.rm=TRUE)
  if(r>0){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rainbow(n+1)[r+1])
  segments(x0=r/(n),y0=0.9,y1=1)
  text(x=r/(n),y=0.8,r)
}
}
plotter()
pdf(file="ninelikes.pdf")
plotter()
dev.off()
```

@@

```{r calculatelikes,cache=F}
a <- seq(from=0.15,to=0.5,by=0.01)
La <- f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)
La <- La - max(La,na.rm=TRUE)
```

@@

```{r calculatelikes1,cache=F}
a <- seq(from=0.005,to=0.6,by=0.005)
La <- f_vec_arn(a,r=results$place[1]-1,results$runners[1]-1,log=TRUE)
La <- La - max(La,na.rm=TRUE)
plot(a,La,type='b')
abline(h=c(0,-2))
```

@@

```{r,label=calcbothlikes,cache=F}
x <- seq(from=0.1,to=0.8,by=0.01)
jj <- as.matrix(expand.grid(x,x))
L <- apply(jj,1,supp2,results$place,results$runners)
L <- L - max(L,na.rm=TRUE)
L <- matrix(L,length(x),length(x))
```

@@

```{r,cache=F}
x <- seq(from=0.1,to=0.95,by=0.01)
jj <- as.matrix(expand.grid(x,x))
L <- apply(jj,1,supp2_ed,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)
L <- L - max(L,na.rm=TRUE)
L <- matrix(L,length(x),length(x))
```
